---
title: "LFC 2022 floristic quality analysis"
format: html
editor: visual
---

```{r preliminaries, include = FALSE}
library(tidyverse)
library(fqar)
library(here)
library(readxl)
library(broom)
library(knitr)
theme_set(theme_minimal())

file1 <- here("lfc_fqa_team1.xlsx")
file2 <- here("lfc_fqa_team2.xlsx")
file3 <- here("lfc_fqa_team3.xlsx")

team1_raw <- read_excel(file1)
team2_raw <- read_excel(file2)
team3_raw <- read_excel(file3)

es224 <- rbind(team1_raw, team2_raw, team3_raw)
es224 <- mutate(es224,
                team = as.factor(team),
                plot_number = as.factor(plot_number),
                species_name = tolower(species_name),
                transect = tolower(transect))
species <- es224 |>
  arrange(species_name) |>
  select(species_name, C) |>
  group_by(species_name, C) |>
  summarize(total_quadrats_found = n())
```

## Species inventory

Over 100 species were observed, in total. The first few are shown below, while the full inventory can be found in the file *species.csv*. Disclaimer: while this data frame include 134 rows (species), some of these are duplicates due to spelling errors.

```{r, echo = FALSE}
kable(species[1:10, ])
```

The breakdown by teams was relatively uniform.

```{r, echo = FALSE}
es224 |> count(team)
ggplot(es224, aes(x = team,
                  fill = team)) +
  geom_bar() +
  theme(legend.position = "none") +
  scale_fill_brewer(palette = "Dark2")
chisq.test(c(84, 94, 72))
```

The p-value of .23 indicates that differences between teams could eaasily be attributed to random chance.

## Transect comparison

The wetland showed both the greatest richness (total number of species observed) and highest native mean-C.

```{r, echo = FALSE, warning = FALSE}
es224_native <- mutate(es224, C = as.numeric(C))

es224_native |>
  group_by(transect) |>
  summarize(species_count = length(unique(species_name)),
            mean_c = mean(C, na.rm = TRUE),
            sd_c = sd(C, na.rm = TRUE)
            ) |>
  arrange(-species_count)
```

The simplicity of this table of numerical summaries is both beautiful and dangerous. A complicated data set like this just can't be reduced to 12-number summary without significant information loss. A picture is a step in the right direction.

```{r, echo = FALSE, warning = FALSE}
ggplot(es224_native, aes(x = transect,
                         y = C,
                         col = transect)) +
  geom_boxplot() +
  geom_jitter(width = .2) +
  theme(legend.position = "none") +
  scale_color_brewer(palette = "Dark2")
```

## Co-occurrence of native species.

Perhaps we can learn something about individual plant species by looking at the C-values of the species they're co-located with, that is, the ones nearby them. For each species, I computed the average C-value of native plants found in quadrats where they were located. The first few rows are shown below. The entire data set is contained in the *colocation.csv* file.

```{r, include = FALSE}
es224_working <- mutate(es224,
                        nativity = case_when(C == "nn" ~ "non-native",
                                              TRUE ~ "native"),
                        C = as.numeric(C))

es224_working <- es224_working |>
  select(transect:plot_number, species_name:C, nativity) |>
  group_by(transect, plot_number) |>
  mutate(local_average_c = mean(C, na.rm = TRUE),
         local_n = sum(!is.na(C))) |>
  ungroup()

cospecies_summary <- es224_working |>
  group_by(species_name, C) |>
  summarize(cospecies_average_c = round(sum(local_average_c * local_n) / sum(local_n), 1),
            cooccurrences = sum(local_n)) |>
  ungroup() |>
  mutate(discrepancy = C - cospecies_average_c)
```

```{r, echo = FALSE}
kable(cospecies_summary[1:10, ])
```

As you'd expect, there is a moderately strong correlation (0.54) between a plant's C-value and the average C-values of those species found nearby.

```{r, echo = FALSE, warning = FALSE, message = FALSE}
ggplot(cospecies_summary, aes(x = C, y = cospecies_average_c)) +
  geom_jitter(width = .2) +
  geom_smooth(method = "lm",
              se = FALSE)

```

## Species profiles

For each species, a profile of co-located C-values can be constructed. The calico aster (*symphyotrichum lateriflorum*, C=4), for instance, breaks down as follows:

```{r, include = FALSE}
species <- "symphyotrichum lateriflorum"

es224_species_quadrats <- filter(es224_working, species_name == species)
target_c <- es224_species_quadrats$C[1] # just looks at first entry. Could be an issue if a species has multiple C values

es224_co <- filter(es224, 0==1) # initialize df with colnames but no rows
for (i in 1:nrow(es224_species_quadrats)){
  temp <- es224 |>
    filter(transect == es224_species_quadrats$transect[i],
           team == es224_species_quadrats$team[i],
           plot_number == es224_species_quadrats$plot_number[i])
  es224_co <- rbind(es224_co, temp)
}

es224_cospecies <- es224_co |>
  mutate(target_species = species,
         target_species_c = target_c,
         C = as.numeric(C)) |>
  select(target_species,
         target_species_c,
         cospecies = species_name,
         cospecies_c = C,
         transect,
         team,
         plot_number) |>
  filter(cospecies != target_species)
```

```{r, echo = FALSE, warning = FALSE}
ggplot(es224_cospecies, aes(x = cospecies_c)) +
  geom_bar() +
  scale_x_continuous(breaks = seq(from = 0, to = 11, by = 2)) +
  labs(x = "Co-occurring species C values",
       y = "Frequency",
       title = "Calico aster native co-occurrence profile") +
  theme_minimal()
```

The profile of the Canada mayflower (*maianthemum canadense*, C=7) is somewhat different:

```{r, include = FALSE}
species <- "maianthemum canadense"

es224_species_quadrats <- filter(es224_working, species_name == species)
target_c <- es224_species_quadrats$C[1] # just looks at first entry. Could be an issue if a species has multiple C values

es224_co <- filter(es224, 0==1) # initialize df with colnames but no rows
for (i in 1:nrow(es224_species_quadrats)){
  temp <- es224 |>
    filter(transect == es224_species_quadrats$transect[i],
           team == es224_species_quadrats$team[i],
           plot_number == es224_species_quadrats$plot_number[i])
  es224_co <- rbind(es224_co, temp)
}

es224_cospecies <- es224_co |>
  mutate(target_species = species,
         target_species_c = target_c,
         C = as.numeric(C)) |>
  select(target_species,
         target_species_c,
         cospecies = species_name,
         cospecies_c = C,
         transect,
         team,
         plot_number) |>
  filter(cospecies != target_species)
```

```{r, echo = FALSE, warning = FALSE}
ggplot(es224_cospecies, aes(x = cospecies_c)) +
  geom_bar() +
  scale_x_continuous(breaks = seq(from = 0, to = 11, by = 2)) +
  labs(x = "Co-occurring species C values",
       y = "Frequency",
       title = "Canada mayflower native co-occurrence profile") +
  theme_minimal()
```

Our sample is very small and specific to our hyper-local environment. Can this idea be scaled up?

## The Chicago database

The website [www.universalfqa.org](www.universalfqa.org) includes hundreds of public floristic quality assessments in standardized format. I downloaded all of the ones that used the most up-to-date database for the Chicagoland area and have begun a co-occurrence analysis. There are over 2.4 million co-occurrences in this set; the full data set is not included, but a summary (*chicago_cooccurrences_summary.csv*) is. The first few rows of this are:

```{r, include = FALSE}
file <- here("chicago_cooccurrences_summary.csv")
chicago_sum <- read_csv(file)
```

```{r, echo = FALSE}
kable(chicago_sum[1:10, ])
```

The species profile for the calico aster is:

```{r, include = FALSE}
file <- here("aster_cooccurrence.csv")
aster_cooccur <- read_csv(file) |> 
  filter(cospecies_nativity == "native")
```

```{r, echo = FALSE, message = FALSE}
ggplot(aster_cooccur, aes(x = cospecies_c)) +
  geom_bar() +
  scale_x_continuous(breaks = seq(from = 0, to = 11, by = 2)) +
  labs(x = "Co-occurring species C values",
       y = "Frequency",
       title = "Calico aster native co-occurrence profile") +
  theme_minimal()

```

A few sample questions that a complete co-occurrence analysis might address:

-   Some species show high discrepancy, co-occuring predominately with species of very different C-values. How might such species be re-classified?

-   Are some species more informative than others?

-   Would it be useful to assign non-native plants C-values on the basis of co-occurring C's?

-   Suppose we used an algorithm to assign new C values on the basis of co-occurrence. What might the output look like? How would those values fare when applied to existing floristic quality assessments?
